---
title: "R Notebook"
output: html_notebook
---

1. As in your previous assignments, compare the accuracy of at least two recommender system algorithms against your offline data.

2. Implement support for at least one business or user experience goal such as increased serendipity, novelty, or diversity.

3. Compare and report on any change in accuracy before and after you’ve made the change in #2.

4. As part of your textual conclusion, discuss one or more additional experiments that could be performed and/or metrics that could be evaluated only if online evaluation was possible. Also, briefly propose how you would design a reasonable online evaluation environment.


```{r}
library(recommenderlab)
library(tidyverse)
```


### Deliverable #1
#### As in your previous assignments, compare the accuracy of at least two recommender system algorithms against your offline data.


```{r}
data("Jester5k")
```

```{r}
dim(Jester5k)
```

```{r}
typeof(Jester5k)
```
All of the objects in `recommenderLab` are created under the [S4 Object Oriented system](http://adv-r.had.co.nz/S4.html), which presents a different approach to thinking about how the dataset is manipulated and how models are used.

```{r}
# Summary of ratings data
jester_df <- as(Jester5k, 'data.frame')
summary(jester_df$rating) 
```
```{r}
# Total number of ratings
nratings(Jester5k)
```
```{r}
# Sparsity percentage
nratings(Jester5k) / (dim(Jester5k)[1] * dim(Jester5k[2]))
```
```{r}
jester_matrix <- spread(jester_df, key = "item", value = "rating")
jester_matrix
```

```{r}
hist(getRatings(Jester5k), breaks=100)
```
We can see when we plot a histogram that shows the negative vales occur with similar frequencies and the positive ratgins are more frequent but slope off as you get towards the max rating of 10. 

Let's take a look at the same distribution after normalization.

One by row centering...
```{r}
hist(getRatings(normalize(Jester5k, method="center")), breaks = 100)
```

And the other by Z-score.
```{r}
hist(getRatings(normalize(Jester5k, method="Z-score")), breaks = 100)
```
We can see the peak ratings range from about 0-2.

Lastly we can take a quick look at how many jokes each user rated and the average rating per joke, by taking the row count and column mean, respectively.

```{r}
hist(rowCounts(Jester5k), breaks = 50)
```


```{r}
hist(colMeans(Jester5k), breaks = 20) 
```
```{r}
popular <- Recommender(Jester5k, method = "POPULAR")

```




```{r}
getModel(popular)$topN
```


```{r}
funniest <- which.max(colMeans(Jester5k))
cat(JesterJokes[funniest])

```

```{r}
jokes_rm <- Jester5k[rowCounts(Jester5k) > 50]

min(rowCounts(jokes_rm))
```




The ratings are on a scale of -10 to +10. Since we have a mean rating of 0.85 and a median of 1.46 we can consider that range the average, and certainly not good by joke standards. Jokes often have tough critics so we will consider a rating of 5 a "good" rating, which is at the edge of the 3rd interquartile range.
```{r}
eval_set <- evaluationScheme(data = jokes_rm,
                             method = "split",
                             train = 0.8,
                             given = 30,
                             goodRating = 5)
 eval_set
```


### UBCF models
```{r warning=FALSE}
ubcf_models <- list(
  ubcf_cos_null = list(name = "UBCF", param = list(method = "cosine", normalize = NULL)),
  ubcf_prs_null = list(name = "UBCF", param = list(method = "pearson", normalize = NULL)),
  ubcf_cos_center = list(name = "UBCF", param = list(method = "cosine", normalize = "center")),
  ubcf_prs_center = list(name = "UBCF", param = list(method = "pearson", normalize = "center")),
  ubcf_cos_z = list(name = "UBCF", param = list(method = "cosine", normalize = "Z-score")),
  ubcf_prs_z = list(name = "UBCF", param = list(method = "pearson", normalize = "Z-score"))
)

ubcf_eval_results <- evaluate(x = eval_set, 
                              method = ubcf_models, 
                              n = seq(10, 100, 10)
                              )
                                 
```


```{r}
plot(ubcf_eval_results, "prec/rec", annotate = T, main = "Precision Recall")
title("UBCF Precision-recall")
```

```{r}
plot(ubcf_eval_results, annotate = T) 
title("UBCF ROC curve")
```

### IBCF Models

```{r warning=FALSE}
ibcf_models <- list(
  ibcf_cos_null = list(name = "IBCF", param = list(method = "cosine", normalize = NULL)),
  ibcf_prs_null = list(name = "IBCF", param = list(method = "pearson", normalize = NULL)),
  ibcf_cos_center = list(name = "IBCF", param = list(method = "cosine", normalize = "center")),
  ibcf_prs_center = list(name = "IBCF", param = list(method = "pearson", normalize = "center")),
  ibcf_cos_z = list(name = "IBCF", param = list(method = "cosine", normalize = "Z-score")),
  ibcf_prs_z = list(name = "IBCF", param = list(method = "pearson", normalize = "Z-score"))
)

ibcf_eval_results <- evaluate(x = eval_set, 
                              method = ibcf_models, 
                              n = seq(10, 100, 10)
                              )
```

```{r}
plot(ibcf_eval_results, "prec/rec", annotate = T, main = "Precision Recall")
title("IBCF Precision-recall")
```

```{r}
plot(ibcf_eval_results, annotate = T) 
title("IBCF ROC curve")
```

### RANDOM models

```{r warning=FALSE}
random_models <- list(
  random_cos_null = list(name = "RANDOM", param = list(method = "cosine", normalize = NULL)),
  random_prs_null = list(name = "RANDOM", param = list(method = "pearson", normalize = NULL)),
  random_cos_center = list(name = "RANDOM", param = list(method = "cosine", normalize = "center")),
  random_prs_center = list(name = "RANDOM", param = list(method = "pearson", normalize = "center")),
  random_cos_z = list(name = "RANDOM", param = list(method = "cosine", normalize = "Z-score")),
  random_prs_z = list(name = "RANDOM", param = list(method = "pearson", normalize = "Z-score"))
)

random_eval_results <- evaluate(x = eval_set, 
                              method = random_models, 
                              n = seq(10, 100, 10)
                              )
```

```{r}
plot(random_eval_results, "prec/rec", annotate = T, main = "Precision Recall")
title("RANDOM Precision-recall")
```
```{r}
plot(random_eval_results, annotate = T) 
title("RANDOM ROC curve")
```

```{r}
popular_models <- list(
  popular_cos_null = list(name = "POPULAR", param = list(method = "cosine", normalize = NULL)),
  popular_prs_null = list(name = "POPULAR", param = list(method = "pearson", normalize = NULL)),
  popular_cos_center = list(name = "POPULAR", param = list(method = "cosine", normalize = "center")),
  popular_prs_center = list(name = "POPULAR", param = list(method = "pearson", normalize = "center")),
  popular_cos_z = list(name = "POPULAR", param = list(method = "cosine", normalize = "Z-score")),
  popular_prs_z = list(name = "POPULAR", param = list(method = "pearson", normalize = "Z-score"))
)
```

```{r}
popular_eval_results <- suppressWarnings(evaluate(x = eval_set, method = popular_models, n = seq(10, 100, 10)))
```


```{r}
plot(popular_eval_results, "prec/rec", annotate = T, main = "Precision Recall")
title("POPULAR Precision-recall")
```

```{r}
plot(popular_eval_results, annotate = T) 
title("POPULAR ROC curve")
```


### Model Selection

After looking at the ROC curves on the four methods, it appears that the IBCF

```{r}
```


```{r}
```

```{r}
# Set the training, known and unknown sets
training_set <- getData(eval_set, "train")

known_set <- getData(eval_set, "known")

unknown_set <- getData(eval_set, "unknown")
```


```{r}

ubcf_rec <- Recommender(data = training_set, method = "UBCF")

ibcf_rec <- Recommender(data = training_set, method = "IBCF")

popular_rec <- Recommender(data = training_set, method = "POPULAR")

random_rec <- Recommender(data = training_set, method = "RANDOM")
```


```{r}
ubcf_model <- predict(ubcf_rec, known_set, type = "ratingMatrix")

ibcf_model <- predict(ibcf_rec, known_set, type = "ratingMatrix")

popular_model <- predict(popular_rec, known_set, type = "ratingMatrix")

random_model <- predict(random_rec, known_set, type = "ratingMatrix")

```

```{r}
ubcf_prs_z_rec <- Recommender(data = training_set, method = "UBCF", parameter = list(method = "pearson", normalize = "Z-score"))

ibcf_prs_c_rec <- Recommender(data = training_set, method = "IBCF", parameter = list(method = "pearson", normalize = "center"))

popular_cos_z_rec <- Recommender(data = training_set, method = "POPULAR", parameter = list(method = "cosine", normalize = "Z-score"))

random_prs_n_rec <- Recommender(data = training_set, method = "RANDOM", parameter = list(method = "pearson", normalize = NULL))
```

```{r}
ubcf_prs_z_model <- predict(ubcf_prs_z_rec, known_set, type = "ratingMatrix")

ibcf_prs_c_model <- predict(ibcf_prs_c_rec, known_set, type = "ratingMatrix")

popular_cos_z_model <- predict(popular_cos_z_rec, known_set, type = "ratingMatrix")

random_prs_n_model <- predict(random_prs_n_rec, known_set, type = "ratingMatrix")

```

```{r}
error <- rbind(
  UBCF = calcPredictionAccuracy(ubcf_model, unknown_set),
  UBCF_prs_z = calcPredictionAccuracy(ubcf_prs_z_model, unknown_set),
  IBCF = calcPredictionAccuracy(ibcf_model, unknown_set),
  IBCF_prs_c = calcPredictionAccuracy(ibcf_prs_c_model, unknown_set),
  POPULAR = calcPredictionAccuracy(popular_model, unknown_set),
  POPULAR_cos_z = calcPredictionAccuracy(popular_cos_z_model, unknown_set),
  RANDOM = calcPredictionAccuracy(random_model, unknown_set),
  RANDOM_prs_n = calcPredictionAccuracy(random_prs_n_model, unknown_set)
)
error
```





















